---
title: "Reproducible Research: Peer Assessment 1"
author: "Alfonso Montero"
date: "15/10/2015"
output: 
  html_document:
    keep_md: true
---

## Loading necesary libraries
```{r, message=FALSE}
library(lubridate)
library(dplyr)
```
## Loading and preprocessing the data
```{r}
unzip("activity.zip","activity.csv")
activity<-read.csv("activity.csv")
```
Convert **date** to a more useful format  
```{r}
activity$date<-ymd(activity$date)
```
## What is mean total number of steps taken per day?
1 Sum of steps taken per day
```{r}
totalStepsByDay <- select(activity, steps, date) %>%
        group_by(date) %>%
        summarise(totalSteps = sum(steps))
```
2 Histogram 
```{r}
plot(totalStepsByDay$date,totalStepsByDay$totalSteps,type="h",
     xlab="Date", ylab="Number of steps", main="Number of steps taken each day")
```

3 Calculate mean and median of steps per day
```{r}
meanSteps<-mean(totalStepsByDay$totalSteps,na.rm=TRUE)
medianSteps<-median(totalStepsByDay$totalSteps,na.rm=TRUE)
cat('Mean:',meanSteps,' Median:',medianSteps)
```

## What is the average daily activity pattern?
1 Summarize the number of steps taken, grouped by intervals
```{r}
totalSteps<-sum(activity$steps,na.rm=TRUE)
totalSteps5min<-select(activity, steps,interval) %>%
        group_by(interval) %>%
        summarise(steps=mean(steps,na.rm=TRUE))
```
2 Plot 
```{r}
plot(totalSteps5min$interval,totalSteps5min$steps,type='l',
     xlab="Interval", ylab="Average steps",main="Average number of steps per Interval")
```

3 The interval with the higher number of steps was:
```{r}
activity[max(activity$steps,na.rm=TRUE),"interval"]
```
## Imputing missing values
1 Total number of rows with missing values in the dataset
```{r}
nrow(activity)-sum(complete.cases(activity))
```
2 Filling NA's with the median of the 5-min interval. 

Calculate the mean for 5-minute intervals.
```{r}
meanSteps5min<-select(activity, steps,interval) %>%
        group_by(interval) %>%
        summarise(meanSteps=mean(steps,na.rm=TRUE))
```
3 Create a dataset with NA's filled
```{r}
activityFilled<-activity
NAindex<-which(is.na(activityFilled$steps))
for (f in NAindex){
        actualInterval<-activityFilled$interval[f]
        activityFilled$steps[f]=meanSteps5min$meanSteps[which(meanSteps5min$interval == actualInterval)]
}
```
Sum of steps taken per day
```{r}
totalStepsByDayFilled <- select(activityFilled, steps, date) %>%
        group_by(date) %>%
        summarise(totalSteps = sum(steps))
```
4 Histogram with NA's filled
```{r}
plot(totalStepsByDayFilled$date,totalStepsByDayFilled$totalSteps,type="h",
     xlab="Date", ylab="Number of steps", main="Number of steps taken each day without NA's")
```

Calculate mean and median of steps per day  
```{r}
meanStepsFilled<-mean(totalStepsByDayFilled$totalSteps,na.rm=TRUE)
medianStepsFilled<-median(totalStepsByDayFilled$totalSteps,na.rm=TRUE)
cat('Mean:',meanStepsFilled,' Median:',medianStepsFilled)
```


## Are there differences in activity patterns between weekdays and weekends?  
1 Create a **factor** column indicating if date is a weekend or not.  
```{r}
weekDay<-function(date){
        nday<-wday(date)
        #1=Sunday, 7=Saturday
        if((nday==1) | (nday==7)) {'weekend'}
        else {'weekday'}
}
activityFilled$weekday<-as.factor(mapply(weekDay,activityFilled$date))
```
2 Make a plot
```{r}
meanSteps5minWeekend<-select(activityFilled, steps,interval,weekday) %>%
        filter(weekday=='weekend') %>%
        group_by(interval) %>%
        summarise(meanSteps=mean(steps,na.rm=TRUE))
meanSteps5minWeekday<-select(activityFilled, steps,interval,weekday) %>%
        filter(weekday=='weekday') %>%
        group_by(interval) %>%
        summarise(meanSteps=mean(steps,na.rm=TRUE))
par(mfrow=c(2,1),mar=c(4,4,2,2),oma=c(2,0,2,0))
plot(meanSteps5minWeekday$interval,meanSteps5minWeekday$meanSteps,type="l",col="blue",
     main="Weekday",xlab="",ylab="Number of steps")
plot(meanSteps5minWeekend$interval,meanSteps5minWeekend$meanSteps,type="l",col="red",
     main="Weekend",xlab="Interval",ylab="Number of steps")

```